---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidytuesdayR)

library(data.table)
library(viridis)
library(tidyr)
library(dplyr)
library(ggplot2)
library(scales)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
df = read.csv("expeditions.csv")
df = data.table(df)

top_15_peaks = df %>%
  select(peak_name, expedition_id) %>%
  count(peak_name, sort = TRUE) %>%
  slice_head(n = 15) %>%
  select(peak_name)

top_15_peaks = df %>%
  inner_join(top_15_peaks, by = "peak_name") %>%
  count(peak_name, season, sort = TRUE)

ggplot(top_15_peaks, aes(x = n, y = reorder(peak_name, n, sum), fill = season)) +
  geom_col() +
  labs(
    title = "The 15 most popular peaks stacked by season of expedition",
    y = NULL,
    x = "Number of expeditions"
    ) +
  theme_light() +
  scale_fill_viridis_d(option = "viridis") +
  theme(legend.position = "bottom")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
df = read.csv("phd_by_field.csv")
df = data.table(df)

phds = df %>%
  group_by(broad_field, year) %>%
  summarize(n = sum(n_phds, na.rm = TRUE), .groups = "drop")

ggplot(phds, aes(x = year, y = n, color = broad_field)) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_color_brewer(palette = "Dark2") +
  theme_light() +
  labs(
    title = "Number of awarded Ph.D.-s in the US by year",
    y = NULL,
    x = NULL,
    color = "Broad field"
    )
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r, warning=FALSE}
df = read.csv("commute.csv")
df = data.table(df)

commute = df %>%
  select(state_region, state_abb, mode, n) %>%
  group_by(state_region, state_abb, mode) %>%
  summarize(N = sum(n, na.rm = TRUE), .groups = "drop") %>%
  filter(mode %in% c("Walk", "Bike")) %>%
  pivot_wider(names_from = mode, values_from = N)

ggplot(commute, aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2) +
  geom_text(aes(label = state_abb), color = "black") +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format(), limits = c(25, NA)) +  # the choice of 25 as starting is arbitrary until the plot looked right :)
  theme_light() +
  labs(
    title = "Number of people walking vs. biking to work in each USA state",
    x = "Number of ppl walking to work (log N)",
    y = "Number of ppl biking to work (log N)",
    color = "State region"
  )

# Note that I don't know why not all state abbreviations appear on the example chart. My result looks therefore a bit messy. :)
```
